Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/introduction.tex
minor rephrasings in intro

Such array computations are prone to error in their indexing
terms. For example, a logical off-by-one-error might manifest itself
as writing $\texttt{a(i)}$ instead of $\texttt{a(i-1)}$ (we revisit an
example we found of this in \Cref{sec:case-studies}).
Errors also arise by simple lexical mistakes when large amounts of
fine-grained indexing are involved in a single expression. For
example, the following snippet from a Navier-Stokes
fluid model~\citet{griebel1997numerical} has two arrays are
read with different data access patterns, across two dimensions, with
dense index-manipulation code:
%%
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) /(4.0*delx)

@@

To support correct array programming, we propose a simple,
abstract specification language for the data access pattern of
array-loop computations. This provides a way to rule out bugs due to
indexing errors. The language design is informed by an initial
empirical study of array computations in a corpus of a real-world
scientific code base, totaling 1.4 million lines of
code (\Cref{sec:study}). We confirm our initial hypotheses of the
ubiquity of looped array computations, but also that they have a
common form, reading from arrays in a fixed neighbourhood of
contiguous elements with a simple static pattern. From this we
designed a simple set of combinators to express common array patterns (\Cref{sec:lang}).
As an example, the Laplace program is specified in our language by:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
That is, \texttt{a} is accessed with a symmetrical pattern in its
first dimension (``centered'') to a depth of one in each direction and
its result contributes to an array
write.  The Navier-Stokes example has two specifications:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2) + centered(depth=1,dim=2)*pointed(dim=1) :: u
!= stencil forward(depth=1,dim=1)*backward(depth=1,dim=2) :: v
\end{minted}
%
The specifications requires that \texttt{u}
is accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v} is
accessed in a neighbourhood bounded forwards to depth of $1$ in the
first dimension and backward to a depth of $1$ in the second
dimension. The specification is relatively small and abstract compared
with the code, with a small number of combinators that \emph{do not
  involve any indexing expressions}, \eg{} \texttt{a(i+1,j-1)}. This
contrasts with other specification approaches, \eg{}, deductive
verification tools such as ACSL~\citet{baudin2008acsl}, where
specifications about array computations must also be expressed in
terms of array-indexing expressions. Thus, any low-level mistakes that
could be made whilst programming complex indexing code could also be
made when writing its specification. Our specifications are
abstract to avoid the error-prone nature of index manipulation; our
specifications are more lightweight to aid its adoption in science.

We implemented a verification tool for our specification language as
an extension to CamFort~\cite{camfort}, an open-source program
analysis tool for Fortran. Specifications are associated to code via
comments, against which the tool checks the code for conformance. This
specify-and-check approach reduces testing effort and future-proofs
against bugs introduced during refactoring and maintenance. A
specification also concisely captures the array access pattern
and provides documentation. We also provide an inference procedure
which efficiently produces specifications with no programmer
intervention, automatically inserting specifications at appropriate
places in the source code. This aids adoption of our approach to
existing legacy code base.

@@

We fix some terminology for syntactic constructs in our target language.

>>>>

Such array computations are prone to error in their indexing
terms. For example, a logical off-by-one-error might manifest itself
as writing $\texttt{a(i)}$ instead of $\texttt{a(i-1)}$ (we revisit an
example we found of this in \Cref{sec:case-studies}).
Errors also arise by simple lexical mistakes when large amounts of
fine-grained indexing are involved in a single expression. For
example, the following snippet from a Navier-Stokes
fluid model~\cite{griebel1997numerical} has two arrays which are
read with different data access patterns, across two dimensions, with
dense index-manipulation code:
%%
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) /(4.0*delx)

@@

To support correct array programming, we propose a simple,
abstract specification language for the data access pattern of
array-loop computations. This provides a way to prevent
indexing errors and also to capture some of the original high-level intent of the algorithm. The language design is informed by an initial
empirical study of array computations in a corpus of a real-world
scientific code base, totaling 1.4 million lines of
code (\Cref{sec:study}). We confirm our initial hypotheses of the
ubiquity of looped array computations, but also that they have a
common form, reading from arrays in a fixed neighbourhood of
contiguous elements with a simple static pattern. From this we
designed a simple set of combinators to express common array patterns (\Cref{sec:lang}).
As an example, the Laplace program is specified in our language by:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
That is, \texttt{a} is accessed with a symmetrical pattern in its
first dimension (``centered'') to a depth of one in each direction and
its result contributes to an array
write.  The Navier-Stokes example has two specifications:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2) + centered(depth=1,dim=2)*pointed(dim=1) :: u
!= stencil forward(depth=1,dim=1)*backward(depth=1,dim=2) :: v
\end{minted}
%
These specifications record that \texttt{u}
is accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v} is
accessed in a neighbourhood bounded forwards to depth of $1$ in the
first dimension and backward to a depth of $1$ in the second
dimension. The specification is relatively small and abstract compared
with the code, with a small number of combinators that \emph{do not
  involve any indexing expressions}, \eg{} \texttt{a(i+1,j-1)}. This
contrasts with other specification approaches, \eg{}, deductive
verification tools such as ACSL~\citet{baudin2008acsl}, where
specifications about array computations must also be expressed in
terms of array-indexing expressions. Thus, any low-level mistakes that
could be made whilst programming complex indexing code could also be
made when writing its specification. Our specifications are
abstract to avoid the error-prone nature of index manipulation; our
specifications are more lightweight to aid its adoption in science.

We implemented a verification tool for our specification language as
an extension to CamFort~\cite{camfort}, an open-source program
analysis tool for Fortran. Specifications are associated with code via
comments, against which the tool checks the code for conformance. This
specify-and-check approach reduces testing effort and future-proofs
against bugs introduced during refactoring and maintenance. A
specification also concisely captures the array access pattern
and provides documentation. We also provide an inference procedure
which efficiently produces specifications with no programmer
intervention, automatically inserting specifications at appropriate
places in the source code. This aids adoption of our approach to
existing legacy code base.

@@

In the remainder of this paper we apply the following terminology for syntactic constructs in our target language.
